---
title: "R Notebook"
output: html_notebook
---

### The project

We will build and implement a trend trading strategy based on the moving average price of the SP500. This is a simple strategy that we will use to explore R code.

A simple moving average is calculated by adding all the data for a number of periods and dividing the total by the number of periods. An exponential moving average assigns greater weight to recent time periods. 

We will be examining a trading strategy that buys when the SP500 50-day SMA is above the 200-day SMA (a 'golden cross') and sells when the 50-day SMA moves below the 200-day SMA (a 'death cross').  

### Load the packages

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(tidyverse)
library(tidyquant)
library(timetk)
library(tibbletime)
library(highcharter)
```

### Import data
We will be working with SP500 and treasury bills data so that when we exit the SP500 we can invest in treasuries and get some yield.

Let's explore two ways to get this date. 

First, it is publicly available. We can use the `tidyquant` package and it's `tq_get()` function to grab the data from yahoo! finance. 

```{r}

symbols <- c("^GSPC", "^IRX")


prices <- 
  tq_get(symbols, 
         get = "stock.prices",
         from = "1980-01-01")

library(writexl)
# Write the first data set in a new workbook
write_xlsx(prices, "prices.xlsx")
write_csv(prices, "prices.csv", append = FALSE)
```


That works well but in industry it's likely you wouldn't be accessing the yahoo! for your price data. Let's see how to import this data from an Excel file or a csv file. 

```{r}
prices_excel <-  
 read_excel("prices.xlsx") %>% 
  mutate(date = ymd(date))
```

```{r}
prices_csv <- 
 read_csv("prices.csv")  %>% 
  mutate(date = ymd(date))
```


### Explore the raw data

Start with the simple line chart. This is the SP500, so we probably won't find much of interest here, but if this were alternative data who knows. 

We will use the `ggplot2` package for this.`ggplot2` is R's most popular data visualization package and we will explore its grammar and layering logic. For now, a line chart.

```{r}
prices %>% 
  dplyr::filter(symbol == "^GSPC") %>% 
  select(date, adjusted) %>% 
  ggplot(aes(x = date, y = adjusted)) +
  geom_line(color = "cornflowerblue")
```

Let's add daily returns and better column names.

```{r}
prices %>% 
  select(symbol, date, adjusted) %>% 
  spread(symbol, adjusted) %>%
  rename(spy = `^GSPC`, treas = `^IRX`) %>% 
  mutate(spy_returns = log(spy) - log(lag(spy))) %>% 
  tail(6)
```


```{r, warning = FALSE}
prices %>% 
  select(symbol, date, adjusted) %>% 
  spread(symbol, adjusted) %>%
  rename(spy = `^GSPC`, treas = `^IRX`) %>% 
  mutate(spy_returns = log(spy) - log(lag(spy))) %>% 
  ggplot(aes(x = date, y = spy_returns)) +
  geom_point(color = "cornflowerblue") +
  scale_x_date(breaks = scales::pretty_breaks(n = 30)) +
  ylab("SP500 daily returns") +
  theme(axis.text.x = element_text(angle = 90))
```

It's a blunt instrument but look at late 1987, 2000-2002 and late 2008 to 2009.

Let's look at how these returns are distributed. 

```{r}
prices %>% 
  select(symbol, date, adjusted) %>% 
  spread(symbol, adjusted) %>%
  rename(spy = `^GSPC`, treas = `^IRX`) %>% 
  mutate(spy_returns = log(spy) - log(lag(spy))) %>% 
  ggplot(aes(x = spy_returns)) +
  geom_histogram(color = "cornflowerblue", binwidth = .003, fill = "pink") +
  # scale_x_date(breaks = scales::pretty_breaks(n = 30)) +
  ylab("SP500 daily returns") +
  theme(axis.text.x = element_text(angle = 90))
```

Eyeball test indicates a longer, negative tail. 

`ggplot()` works in layers. We can add a density layer. 

```{r}
prices %>% 
  select(symbol, date, adjusted) %>% 
  spread(symbol, adjusted) %>%
  rename(spy = `^GSPC`, treas = `^IRX`) %>% 
  mutate(spy_returns = log(spy) - log(lag(spy))) %>% 
  ggplot(aes(x = spy_returns)) +
  geom_histogram(color = "cornflowerblue", binwidth = .003, fill = "pink") +
  stat_density(geom = "line", color = "green") +
  ylab("SP500 daily returns") +
  theme(axis.text.x = element_text(angle = 90))
```

Let's investigate the tail statistics a little bit.

```{r}
prices %>% 
  select(symbol, date, adjusted) %>% 
  spread(symbol, adjusted) %>%
  rename(spy = `^GSPC`, treas = `^IRX`) %>% 
  mutate(spy_returns = log(spy) - log(lag(spy))) %>%
  na.omit() %>% 
  summarise(skewness = skewness(spy_returns),
            kurtosis = kurtosis(spy_returns))
```

Negative skew and excess kurtosis; our returns are not exactly normally distributed. 

We have a feel for our data, let's start to add our logic and transform this data. 

First, we need a way to calculate the rolling 50 and 200-day moving averages. We use `rollify()` to create functions for this.


```{r}

roll_mean_50 <- 
  rollify(mean, window = 50)

roll_mean_200 <- 
  rollify(mean, window = 200)
```

Now we use `mutate()` to add the moving averages to the data.

```{r}

prices %>% 
  select(symbol, date, adjusted) %>% 
  spread(symbol, adjusted) %>%
  rename(spy = `^GSPC`, treas = `^IRX`) %>% 
  mutate(sma_200 = roll_mean_200(spy),
         sma_50 = roll_mean_50(spy)) %>% 
  na.omit()
```

Let's visualize our new trends, compared to raw price. 

```{r}

prices %>% 
  select(symbol, date, adjusted) %>% 
  spread(symbol, adjusted) %>%
  rename(spy = `^GSPC`, treas = `^IRX`) %>% 
  mutate(sma_200 = roll_mean_200(spy),
         sma_50 = roll_mean_50(spy)) %>% 
  na.omit() %>% 
  select(-treas) %>% # -spy to see just the trends
  dplyr::filter(date > "2015-01-01") %>% # try different date bands, post 2017, 2018
  gather(series, value, -date)  %>% 
  ggplot(aes(x = date, y = value, color = series)) +
  geom_line()
```

Let's get algorithmic.

*If* the 50-day MA is above the 200-day MA, buy the market, *else* go to the risk free return, or can put to zero if we prefer.

What we need: 
1) rolling 50-day SMA
2) rolling 200-day SMA
3) if_else logic to create a buy or sell signal

```{r}
prices %>% 
  select(symbol, date, adjusted) %>% 
  spread(symbol, adjusted) %>%
  rename(spy = `^GSPC`, treas = `^IRX`) %>% 
  mutate(sma_200 = roll_mean_200(spy),
         sma_50 = roll_mean_50(spy)) %>% 
  na.omit() %>% 
  mutate(signal = if_else(sma_50 > sma_200, 1, 0),  
         spy_returns = round(log(spy) - log(lag(spy)), 4))  %>% 
  dplyr::filter(date > "1987-10-01" & date < "1988-01-01")
```


Let's add the logic: if the signal is 1, buy the SP500, which we will code as signal * SP500 returns (the next day), else go to risk free.

```{r}

prices %>% 
  select(symbol, date, adjusted) %>% 
  spread(symbol, adjusted) %>%
  rename(spy = `^GSPC`, treas = `^IRX`) %>% 
  mutate(sma_200 = roll_mean_200(spy),
         sma_50 = roll_mean_50(spy),
         signal = if_else(sma_50 > sma_200, 1, 0),
         spy_returns = log(spy) - log(lag(spy)), 
         daily_treas = (1 + (treas/100)) ^ (1/252) - 1,
         trend_returns = if_else(lag(signal) == 1, (signal * spy_returns), daily_treas))
```

We now have a column for our strategy. Let's add a buy and hold strategy where we buy the SP500 90% and treasury rate 10% and hold it for the duration. 

```{r}

prices %>% 
  select(symbol, date, adjusted) %>% 
  spread(symbol, adjusted) %>%
  rename(spy = `^GSPC`, treas = `^IRX`) %>% 
  mutate(sma_200 = roll_mean_200(spy),
         sma_50 = roll_mean_50(spy)) %>% 
  na.omit() %>% 
  mutate(signal = if_else(sma_50 > sma_200, 1, 0),
         spy_returns = log(spy) - log(lag(spy)),
         daily_treas = (1 + (treas/100)) ^ (1/252) - 1,
         buy_hold_returns = (.9 * spy_returns) + (.1 * daily_treas),
         trend_returns = if_else(lag(signal) == 1, (signal * spy_returns), daily_treas)
         ) %>%
  na.omit()
```

We can add columns to see dollar growth for our two asset mixes as well. 

```{r}
sma_trend_results <- 
prices %>% 
  select(symbol, date, adjusted) %>% 
  spread(symbol, adjusted) %>%
  rename(spy = `^GSPC`, treas = `^IRX`) %>% 
  mutate(sma_200 = roll_mean_200(spy),
         sma_50 = roll_mean_50(spy),
         signal = if_else(sma_50 > sma_200, 1, 0),
         spy_returns = log(spy) - log(lag(spy)), 
         daily_treas = (1 + (treas/100)) ^ (1/252) - 1,
         buy_hold_returns = (.9 * spy_returns) + (.1 * daily_treas),
         trend_returns = if_else(lag(signal) == 1, (signal * spy_returns), daily_treas)
         ) %>%
  na.omit() %>% 
  mutate(
         trend_growth = accumulate(1 + trend_returns, `*`),
         buy_hold_growth = accumulate(1 + buy_hold_returns, `*`))

sma_trend_results %>% tail()
```

### Visualize our trend strategy results

```{r}
sma_trend_results %>%
  select(date, trend_growth, buy_hold_growth) %>% 
  gather(strategy, growth, -date) %>% 
  ggplot(aes(x = date, y = growth, color = strategy)) +
  geom_line()

```

Our original trend has grown higher, but the z-score logic seems more stable.

### Analyze with standard deviation, skewness, kurtosis, sharpe ratio


```{r}
rfr <- .00003

sma_trend_results %>%
  select(date, trend_returns, buy_hold_returns) %>% 
  gather(strategy, returns, -date) %>% 
  group_by(strategy) %>% 
  summarise(stddev = sd(returns),
            mean = mean(returns),
            skewnsess = skewness(returns),
            kurtosis = kurtosis(returns),
            # Sharpe ratio with own equation
            sharpe = mean(returns - rfr)/
              sd(returns - rfr))
```

Our strategy has a higher sharpe, and lower standard deviation than buy hold. 
We can plot those for a better sense. 

```{r}

sma_trend_results %>%
  select(date, trend_returns, buy_hold_returns) %>% 
  gather(strategy, returns, -date) %>% 
  group_by(strategy) %>% 
  summarise(stddev = sd(returns),
            mean = mean(returns),
            skewnsess = skewness(returns),
            kurtosis = kurtosis(returns),
            # Sharpe ratio with own equation
            sharpe = mean(returns - rfr)/
              sd(returns - rfr)) %>% 
  ggplot(aes(x = stddev, y = sharpe, color = strategy)) +
  geom_point(size = 3)
```


JKR Note: more visualizations here to show more ggplot functionality. 

The above is a simple implementation, but is flexible. We can change assets, indicators.

We can also add a signal, for example a zscore for when market is a number of standard deviations above or below a rolling average? 
How would we implement that? 

1) Calculate a spread
2) turn it into a z-score
3) create a signal

```{r}
prices %>% 
  select(symbol, date, adjusted) %>% 
  spread(symbol, adjusted) %>%
  rename(spy = `^GSPC`, treas = `^IRX`) %>% 
  mutate(sma_200 = roll_mean_200(spy),
         sma_50 = roll_mean_50(spy)) %>% 
  na.omit() %>% 
  mutate(trend_signal = if_else(sma_50 > sma_200, 1, 0),
         spy_returns = log(spy) - log(lag(spy)),
         z_spread = (spy - sma_200),
         z_score = (z_spread - mean(z_spread))/sd(z_spread),
         z_signal = if_else(
                            lag(z_score, 1) < -.05 & 
                            lag(z_score, 2) < -.05 &
                            lag(z_score, 3) < -.05,
                            #lag(z_score, 4) < -.05 &
                            #lag(z_score, 5) < -.05, 
                            0, 1)) # %>% 
  # dplyr::filter(date > "1987-10-01")
  
```

Now we create a new strategy that depends on both the trend and the z-score.

```{r}
prices %>% 
  select(symbol, date, adjusted) %>% 
  spread(symbol, adjusted) %>%
  rename(spy = `^GSPC`, treas = `^IRX`) %>% 
  mutate(sma_200 = roll_mean_200(spy),
         sma_50 = roll_mean_50(spy),
         spy_returns = log(spy) - log(lag(spy))) %>% 
  na.omit() %>% 
  mutate(trend_signal = if_else(sma_50 > sma_200, 1, 0),
         z_spread = (spy - sma_200),
         z_score = (z_spread - mean(z_spread))/sd(z_spread),
         z_signal = if_else(
                            lag(z_score, 1) < -.05 & 
                            lag(z_score, 2) < -.05 &
                            lag(z_score, 3) < -.05,
                            #lag(z_score, 4) < -.05 &
                            #lag(z_score, 5) < -.05, 
                            0, 1),
         daily_treas = (1 + (treas/100)) ^ (1/252) - 1,
         buy_hold_returns = (.9 * spy_returns) + (.1 * daily_treas),
         trend_z_returns = if_else(lag(trend_signal) == 1 &
                                 z_signal == 1, 
                                 (trend_signal * spy_returns), daily_treas),
         trend_returns =  if_else(lag(trend_signal) == 1,
                                 (trend_signal * spy_returns), daily_treas)) # %>% 
  # dplyr::filter(date > "1987-10-01")
```

Note the NAs - those will be a problem we convert to dollar growth for visualization.

```{r}
trend_z_results <- 
prices %>% 
  select(symbol, date, adjusted) %>% 
  spread(symbol, adjusted) %>%
  rename(spy = `^GSPC`, treas = `^IRX`) %>% 
  mutate(sma_200 = roll_mean_200(spy),
         sma_50 = roll_mean_50(spy),
         spy_returns = log(spy) - log(lag(spy))) %>% 
  na.omit() %>% 
  mutate(trend_signal = if_else(sma_50 > sma_200, 1, 0),
         z_spread = (spy - sma_200),
         z_score = (z_spread - mean(z_spread))/sd(z_spread),
         z_signal = if_else(
                            lag(z_score, 1) < -.05 & 
                            lag(z_score, 2) < -.05 &
                            lag(z_score, 3) < -.05,
                            #lag(z_score, 4) < -.05 &
                            #lag(z_score, 5) < -.05, 
                            0, 1),
         daily_treas = (1 + (treas/100)) ^ (1/252) - 1,
         trend_z_returns = if_else(lag(trend_signal) == 1 &
                                 z_signal == 1, 
                                 (trend_signal * spy_returns), daily_treas),
         trend_returns =  if_else(lag(trend_signal) == 1,
                                 (trend_signal * spy_returns), daily_treas),
         buy_hold_returns = (.9 * spy_returns) + (.1 * daily_treas)) %>% 
  #dplyr::filter(date > "1987-10-01") %>% 
  select(date, trend_signal, z_signal, buy_hold_returns, trend_returns, trend_z_returns) %>%
  na.omit() %>% 
  mutate(
         trend_growth = accumulate(1 + trend_returns, `*`),
         trend_z_growth = accumulate(1 + trend_z_returns, `*`),
         buy_hold_growth = accumulate(1 + buy_hold_returns, `*`))

trend_z_results %>% tail()

```

### Visualize our trend + z strategy results

```{r}
trend_z_results %>%
  select(date, trend_growth, trend_z_growth, buy_hold_growth) %>% 
  gather(strategy, growth, -date) %>% 
  ggplot(aes(x = date, y = growth, color = strategy)) +
  geom_line()

```

Our original trend has grown higher, but the z-score logic seems more stable.

### Analyze with standard deviation, skewness, kurtosis, sharpe ratio


```{r}
rfr <- .00003

trend_z_results %>%
  select(date, trend_returns, trend_z_returns, buy_hold_returns) %>% 
  gather(strategy, returns, -date) %>% 
  group_by(strategy) %>% 
  summarise(stddev = sd(returns),
            mean = mean(returns),
            skewnsess = skewness(returns),
            kurtosis = kurtosis(returns),
            # Sharpe ratio with own equation
            sharpe = mean(returns - rfr)/
              sd(returns - rfr))
```

Our strategies have higher sharpe, and lower standard deviation than buy hold. 
We can plot those for a better sense. 

```{r}

trend_z_results %>%
  select(date, trend_returns, trend_z_returns, buy_hold_returns) %>% 
  gather(strategy, returns, -date) %>% 
  group_by(strategy) %>% 
  summarise(stddev = sd(returns),
            mean = mean(returns),
            skewnsess = skewness(returns),
            kurtosis = kurtosis(returns),
            # Sharpe ratio with own equation
            sharpe = mean(returns - rfr)/
              sd(returns - rfr)) %>% 
  ggplot(aes(x = stddev, y = sharpe, color = strategy)) +
  geom_point(size = 3)
```


### To Shiny so can tweak parameters or RMD for a report
Port this work to Shiny so end user can choose different assets, different moving averages, different z-score levels.

 




